---
title: "Variant calling and blacklist creation (using cohort data)"
output: html_document
vignette: >
  %\VignetteIndexEntry{Variant calling and blacklist creation (using cohort data)"}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

Given single-cell RNA-seq data **from several individuals**, this vignette describes how to identify somatic mitochondrial variants that are specific to single individuals, while at the same time creating a blacklist of sites that are observed in several individuals. The rational for this strategy is that artefactual (RNA-only) variants are often shared between individuals, whereas true somatic variants are not.

## Description of input data

The first step for prepocessing will be to count variants (A, C, G, T, insertion, deletion) at every site of the mitochondrial genome. For this, a list of BAM files is required; to access the BAM files and raw data used for this study, we have to refer to [EGA (Study EGAS00001003414)](https://www.ebi.ac.uk/ega/studies/EGAS00001003414). Assuming bam files are in a folder `bam`, count tables can be created with


```{r createCountTables, eval=F}
countTables <- baseCountsFromBamList( list.files("bam", full.names = T) )
```

Here, we download pre-computed base count tables. We note that these consist of lists of matrices. Each entry in the list corresponds to a single cell. Matrix rows are sites in the mitochondrial genome, and columns are counts 

```{r downloadCountTables}
countTables <- readRDS(url("http://steinmetzlab.embl.de/mutaseq/nuc.count.per.position.RDS"))
countTables <- countTables[grep('Q59',names(countTables),invert=T)]
print(head(countTables[[1]]))
```

Finally, we parse out the patient that each cell in the list derives from. We also count the number of cells per patient.

```{r parsePatient}
patient <- gsub("_.+","",names(countTables))
print(table(patient))
```

## Cohort-level variant calling

We next use the `mutationCallsFromCohort` function to a) filter on coverage to exclude potentially noisy variants and b) compare allele frequencies between patients to remove variants that were observed in several individuals and that therefore are unlikely to represent true somatic variants (e.g. RNA editing events).

```{r pressure}
library(mitoClone)
result <- mutationCallsFromCohort(countTables, patient)
print(colnames(result$P342@M))
print(colnames(result$HRK@M))
```

The result for patient P1 (`result$P342`) an P2 (`result$HRK`) serves as input into the vignette [Computation of phylogenetic trees and clustering of mutations](clustering.html). The blacklist (`result$blacklist`) serves as input into the vignette [Variant calling and validation](calling.html). We have added regions based on a conservative set of false-positive variants from our datsets and further trinucleotide repeats and softmased regions to this blacklist, since we observed that those tend to contain false positives as well. See the `blacklist` object provided as part of the package.
